---
title: "HW4 - DATA622"
author: "Group 1"
date: "11/3/2021"
output:
  html_document: 
    toc: true
    toc-title: "HW4 - DATA622"
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: false
    theme: united
    highlight: tango
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


### Group 1 Members:
-   David Moste
-   Vinayak Kamath
-   Kenan Sooklall
-   Christian Thieme
-   Lidiia Tronina

\pagebreak

## Introduction

For this project, we will be working with a mental health [dataset](https://raw.githubusercontent.com/christianthieme/Machine_Learning_Big_Data/main/HW4/ADHD_data.csv) obtained from a previously completed ADHD research project. The dataset contains information from 175 individuals who took part in the study and includes 54 columns broken into the following categories: 

| Excel Column | Variable                 | Key                                  |
|--------------|--------------------------|--------------------------------------|
| C            | Sex                      |  Male-1, Female-2                    |
| D            | Race                     |  White-1, African American-2, Hispanic-3, Asian-4, Native American-5, Other or missing data -6 |
| E-W          | ADHD self-report scale | Never-0, rarely-1, sometimes-2, often-3, very often-4 |
| X-AM         | Mood disorder questions |  No-0, yes-1; question 3: no problem-0, minor-1, moderate-2, serious-3 |
| AN-AS        | Individual substances misuse |  no use-0, use-1, abuse-2, dependence-3 |
| AT           | Court Order |  No-0, Yes-1 |
| AU           | Education | 1-12 grade, 13+ college |
| AV           | History of Violence | No-0, Yes-1 |
| AW           | Disorderly Conduct | No-0, Yes-1 |
| AX           | Suicide attempt | No-0, Yes-1 |
| AY           | Abuse History | No-0, Physical (P)-1, Sexual (S)-2, Emotional (E)-3, P&S-4, P&E-5, S&E-6, P&S&E-7 |
| AZ           | Non-substance-related Dx | 0 – none; 1 – one; 2 – More than one |
| BA           | Substance-related Dx | 0 – none; 1 – one Substance-related; 2 – two; 3 – three or more |
| BB           | Psychiatric Meds | 0 – none; 1 – one psychotropic med; 2 – more than one psychotropic med | 


Our task is first to explore the data, then using the understanding gained, to build a clustering model to determine different segments of patients. Additionally, we'll perform Principal Component Analysis on a subset of the dataset to see what information can be gleaned. Finally, we'll use Gradient Boosting and Support Vector Machines to predict if a patient will commit suicide.  

As can be seen from the table above, the dataset has a wide variety of numeric and categorical data. We'll dive deeper into this data now.

## Exploratory Data Analysis

```{r message=FALSE, warning=FALSE}
adhd <- readr::read_csv('https://raw.githubusercontent.com/christianthieme/Machine_Learning_Big_Data/main/HW4/ADHD_data.csv')
```



