---
title: "Homework 3"
author: "Group 1"
date: "9/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Homework 3

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(inspectdf)
```


## Load Data

```{r}
loans <- readr::read_csv('https://raw.githubusercontent.com/christianthieme/Machine_Learning_Big_Data/main/HW3/Loan_approval.csv')
```
## Exploratory Data Analysis

To begin our exploratory data analysis, we'll begin by taking a high level look at our dataset: 

```{r}
glimpse(loans)
```
From this view we can see that this dataset has 614 rows and 13 columns. There is a mix of both numeric and categorical data, although it appears that at least one numeric column should actually be categorical. 

```{r}
inspectdf::inspect_types(loans) %>%
  show_plot
```

The data appears to have some missing values that we'll have to deal with as we move along.

Since we know there are missing values, let's see how pervasive the issue is: 

```{r fig.width = 12, fig.height=6}
visdat::vis_miss(loans, sort_miss = TRUE)
```

7 columns have missing data, ranging from 0.49% to 8.14%. What's interesting is there are 22 instances where `LoanAmount` is missing. In total, ~2% of the data has missing values. 

Let's now inspect our numeric features: 

```{r fig.height=5, fig.width=10}
inspectdf::inspect_num(loans) %>% 
 show_plot()
```

Looking at the above plot, we can see that: 

* `ApplicantIncome` must be on a weekly or monthly basis as most values are fairly low
* `CoapplicantIncome` follows the same trend as applicant income
* `Credit_History` appears to be a categorical variable either 0 or 1
* `Loan_Amount_Term` appears to be 365 days (1 year) more than 80% of the time
* `LoanAmount` is fairly small, ranging from \$0-700. It appears that most loans are for somewhere between \$50 and \$200. 

Now, let's turn our attention to our categorical variables. We'll first look to see what percentage each category within a given variable makes up: 

```{r}
inspectdf::inspect_imb(loans %>% select(-Loan_ID))
```

While the table above is insightful, it's also helpful to view it in a visual format: 

```{r}
inspect_cat(loans %>% select(-Loan_ID)) %>% 
  show_plot()
```

Some very interesting points about the demographic we are looking at come as we look at this visual:

* Half of the population do not have any dependents 
* Most people in this dataset are graduates
* Over 75% of the population is male
* 65% of the population is married 
* Most of the individuals are not self-employed


```{r fig.height=4, fig.width=10, message=FALSE, warning=FALSE}
  loan_names <- loans %>% select_if(is.numeric) %>% select(-Credit_History, -Loan_Amount_Term)
  int_names <- names(loan_names)
  myGlist <- vector('list', length(loan_names))
  names(myGlist) <- int_names
  
  for (i in int_names) {       
 
   myGlist[[i]] <- 
       ggplot(loans) + 
       aes_string(x = as.factor(loans$Loan_Status), y = i) + 
       geom_boxplot(color = 'steelblue', 
                    outlier.color = 'firebrick', 
                    outlier.alpha = 0.35) +
        labs(title = paste0(i,' vs Loan_Status'), y = i, x= 'Loan_Status') +
        theme_minimal() + 
        theme(
          plot.title = element_text(hjust = 0.45),
          panel.grid.major.y =  element_line(color = "grey", 
                                             linetype = "dashed"),
          panel.grid.major.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          panel.grid.minor.x = element_blank(),
          axis.ticks.x = element_line(color = "grey")
        )
       
      }

    myGlist <- within(myGlist, rm(target_name))
    gridExtra::grid.arrange(grobs = myGlist, ncol = 3)
```



